SI: Settlements

### 2021-02-11 analysis/paper/paper.Rmd
cont watercourse density

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
#  "This is abstract"
keywords: |
  keyword 1; keyword 2; keyword 3
---

@@

```{r packages}
devtools::load_all(".")
library(tidyverse)
library(sf)
```

@@

* Pottery traditions
* Pottery groups

@@

# read data ---------------------------------------------------------------
# mask
mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)

@@

<!--
```{r base-tab}
# set_base$base %>%
#   DT::datatable(rownames = FALSE)
```
-->

@@

```{r set1-kde}
kde1 <- kdestimate(settlements1,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), kde)
```

@@

```{r set2-kde}
kde2 <- kdestimate(settlements2,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), kde)
```

@@

@@

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```
```{r rm-dist, cache = FALSE}
# calculated separately in a script `rm_distance.R`
# dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
# dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))
```

@@

<!--\@ref(fig:demo-plot)-->

@@

<div id="refs"></div>

>>>>

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

@@

```{r packages}
devtools::load_all(".")
library(sf)
library(tidyverse)
```

@@

-   Pottery traditions
-   Pottery groups

@@

# mask
mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)

@@

@@

```{r set1-kde}
kde1 <- kdestimate(settlements1,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res1_long <- kde1 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

@@

```{r set2-kde}
kde2 <- kdestimate(settlements2,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res2_long <- kde2 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

@@

Minimum, mean, median and maximum distances are calculated, median is used 
as a result.

@@

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```

```{r rm-dist, cache = FALSE}
# calculated separately in a script `rm_distance.R`
# dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
# dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))

rm_dist_median <- rm_dist %>% 
  select(-region, -type, -shp, -min, -mean, -max) %>% 
  group_by(id, rm) %>% 
  summarise(dist_rm = mean(median)) %>% 
  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
  select(-rm)

# update result tibbles
res1_long <- set_base$period1 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res2_long)
```

@@

# Density of watercourses

In settlement structure studies it is often assumed that past human settlements
have some kind of a relation to the network of watercourses. Numerous problems
are faced while assessing the relationship between settlements and close
watercourse. Most importantly the man-made recent changes in the structure and
intensity of watercourses (and landscape in general) are a major factor
imapiring such analysis.

Usually, the settlement - watercourse relationship is expressed in some form of
a metric giving shortest distance or so. We choose an approach used to study
road networks in modern urban areas [e. g. @lin2020].

Most of the input data is derived from *DIBAVOD* data set A01 layer
(<https://www.dibavod.cz/>). 
Data for Slovakia are derived from OpenStreet map ? , data for Austria ???

Density (kernel smoothed intensity) is estimated using `spatstat::density.psp`
function with sigma 2000 m.

```{r get-raster}
hydro_dens <- raster::raster(
  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
  raster::rasterToPoints() %>%
  as_tibble() %>%
  rename(KDE = density_raster)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
ggplot() +
  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void()
```

# Results

```{r results-export}
res1_long %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  arrange(id, period, chrono) %>% 
  write_csv(here::here(derived_data, "results", "results1.csv"))

res2_long %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  arrange(id, period, chrono) %>% 
  write_csv(here::here(derived_data, "results", "results2.csv"))
```

@@

::: {#refs}
:::
